---
title: "In-class_Ex02: Geographical Segmentation with Spatially Constrained Clustering Techniques"
editor: visual
---

# 5.1 Overview

In this hands-on exercise, we will gain hands-on experience on how to delineate homogeneous region by using geographically referenced multivariate data. There are two major analysis, namely:

-   hierarchical cluster analysis; and

-   
# 5.2 Getting Started
## 5.2.1 The analytical question
In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate Shan State, Myanmar into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

# 5.3 The data
Two data sets will be used in this study. They are:

Myanmar Township Boundary Data
Shan-ICT.csv

## 5.3.1 Installing and loading R packages
The code chunks below installs and launches these R packages into R environment.
```{r}
pacman::p_load(rgdal, spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse)

```

# 5.4 Data Import and Prepatation
##  5.4.1 Importing geospatial data into R environment

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

We can view the content of the newly created shan_sf simple features data.frame by using the code chunk below.
```{r}
shan_sf
```

Since shan_sf is conformed to tidy framework, we can also glimpse() to reveal the data type of it’s fields.
```{r}
glimpse(shan_sf)
```

## 5.4.2 Importing aspatial data into R environment
```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")

```

The code chunk below reveal the summary statistics of ict data.frame.
```{r}
summary(ict)
```

## 5.4.3 Derive new variables using dplyr package
```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Let us review the summary statistics of the newly derived penetration rates using the code chunk below.
```{r}
summary(ict_derived)
```

# 5.5 Exploratory Data Analysis (EDA)
Histogram is useful to identify the overall distribution of the data values (i.e. left skew, right skew or normal distribution)
```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Boxplot is useful to detect if there are outliers.
```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Next, we will also plotting the distribution of the newly derived variables (i.e. Radio penetration rate) by using the code chunk below.
```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```


```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")

```

The code chunks below are used to create the data visualisation. They consist of two main parts. First, we will create the individual histograms using the code chunk below.
```{r}

radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Next, the ggarange() function of ggpubr package is used to group these histograms together.
```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

The code chunks below is used to perform the task. The unique identifier used to join both data objects is TS_PCODE.
```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, 
                     by=c("TS_PCODE"="TS_PCODE"))
write_rds(shan_sf, "data/rds/shan_sf.rds")

```

### 5.5.2.2 Preparing a choropleth map
The code chunks below are used to prepare the choroplethby using the qtm() function of tmap package.
```{r}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the total number of households (i.e. TT_HOUSEHOLDS.map) and one for the total number of household with Radio (RADIO.map) by using the code chunk below.
```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate by using the code chunk below.
```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

# 5.6 Correlation Analysis
```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

# 5.7 Hierarchy Cluster Analysis
## 5.7.1 Extrating clustering variables
```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Next, we need to change the rows by township name instead of row number by using the code chunk below
```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Now, we will delete the TS.x field by using the code chunk below.
```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

## 5.7.2 Data Standardisation
In the code chunk below, normalize() of heatmaply package is used to stadardisation the clustering variables by using Min-Max method. The summary() is then used to display the summary statistics of the standardised clustering variables.
```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

## 5.7.4 Z-score standardisation
Z-score standardisation can be performed easily by using scale() of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.
```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

## 5.7.5 Visualising the standardised clustering variables
The code chunk below plot the scaled Radio_PR field.
```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```
What statistical conclusion can you draw from the histograms above?
```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

## 5.7.6 Computing proximity matrix
The code chunk below is used to compute the proximity matrix using euclidean method.
```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

The code chunk below can then be used to list the content of proxmat for visual inspection.
```{r}
proxmat
```

## 5.7.7 Computing hierarchical clustering
The code chunk below performs hierarchical cluster analysis using ward.D method. The hierarchical clustering output is stored in an object of class hclust which describes the tree produced by the clustering process.
```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

We can then plot the tree by using plot() of R Graphics as shown in the code chunk below.
```{r}
plot(hclust_ward, cex = 0.6)
```

## 5.7.8 Selecting the optimal clustering algorithm
The code chunk below will be used to compute the agglomerative coefficients of all hierarchical clustering algorithms.
```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

## 5.7.9 Determining Optimal Clusters
There are three commonly used methods to determine the optimal clusters, they are:

Elbow Method
Average Silhouette Method
Gap Statistic Method
### 5.7.9.1 Gap Statistic Method
To compute the gap statistic, clusGap() of cluster package will be used.
```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Next, we can visualise the plot by using fviz_gap_stat() of factoextra package.
```{r}
fviz_gap_stat(gap_stat)
```

## 5.7.10 Interpreting the dendrograms
It’s also possible to draw the dendrogram with a border around the selected clusters by using rect.hclust() of R stats. The argument border is used to specify the border colors for the rectangles.
```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

## 5.7.11 Visually-driven hierarchical clustering analysis
With heatmaply, we are able to build both highly interactive cluster heatmap or static cluster heatmap.
### 5.7.11.1 Transforming the data frame into a matrix
The code chunk below will be used to transform shan_ict data frame into a data matrix.
```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

### 5.7.11.2 Plotting interactive cluster heatmap using heatmaply()
In the code chunk below, the heatmaply() of heatmaply package is used to build an interactive cluster heatmap.
```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

## 5.7.12 Mapping the clusters formed
With closed examination of the dendragram above, we have decided to retain six clusters.

cutree() of R Base will be used in the code chunk below to derive a 6-cluster model.
```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The code chunk below form the join in three steps:

the groups list object will be converted into a matrix;
cbind() is used to append groups matrix onto shan_sf to produce an output simple feature object called shan_sf_cluster; and
rename of dplyr package is used to rename as.matrix.groups field as CLUSTER.
```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Next, qtm() of tmap package is used to plot the choropleth map showing the cluster formed.
```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

# 5.8 Spatially Constrained Clustering: SKATER approach
## 5.8.1 Converting into SpatialPolygonsDataFrame
```{r}
shan_sp <- as_Spatial(shan_sf)
```

## 5.8.2 Computing Neighbour List
Next, poly2nd() of spdep package will be used to compute the neighbours list from polygon list.
```{r}

shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can plot the neighbours list on shan_sp by using the code chunk below. 
```{r}
plot(shan_sp, 
     border=grey(.5))
plot(shan.nb, 
     coordinates(shan_sp), 
     col="blue", 
     add=TRUE)
```

## 5.8.3 Computing minimum spanning tree
### 5.8.3.1 Calculating edge costs
The code chunk below is used to compute the cost of each edge.
```{r}

lcosts <- nbcosts(shan.nb, shan_ict)
```

Note that we specify the style as B to make sure the cost values are not row-standardised.
```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

## 5.8.4 Computing minimum spanning tree
The minimum spanning tree is computed by mean of the mstree() of spdep package as shown in the code chunk below.
```{r}
shan.mst <- mstree(shan.w)
```

After computing the MST, we can check its class and dimension by using the code chunk below.
```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

We can display the content of shan.mst by using head() as shown in the code chunk below.
```{r}
head(shan.mst)
```

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge.
```{r}
plot(shan_sp, border=gray(.5))
plot.mst(shan.mst, 
         coordinates(shan_sp), 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

## 5.8.5 Computing spatially constrained clusters using SKATER method
The code chunk below compute the spatially constrained cluster using skater() of spdep package.
```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

The result of the skater() is an object of class skater. We can examine its contents by using the code chunk below.
```{r}
str(clust6)
```

We can check the cluster assignment by using the conde chunk below.
```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. 
```{r}
table(ccs6)
```

Lastly, we can also plot the pruned tree that shows the five clusters on top of the townshop area.
```{r}
plot(shan_sp, border=gray(.5))
plot(clust6, 
     coordinates(shan_sp), 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

## 5.8.6 Visualising the clusters in choropleth map
The code chunk below is used to plot the newly derived clusters by using SKATER method.
```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

For easy comparison, it will be better to place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.
```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

# 5.9 Spatially Constrained Clustering: ClustGeo Method
## 5.9.1 Ward-like hierarchical clustering: ClustGeo
To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.
```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

### 5.9.1.1 Mapping the clusters formed
Similarly, we can plot the clusters on a categorical area shaded map by using the steps we learned in 5.7.12 Mapping the clusters formed.
```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### 5.9.2 Spatially Constrained Hierarchical Clustering
Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using st_distance() of sf package.
```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

Next, choicealpha() will be used to determine a suitable value for the mixing parameter alpha as shown in the code chunk below.
```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

With reference to the graphs above, alpha = 0.3 will be used as shown in the code chunk below.
```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.3)
```

Next, cutree() is used to derive the cluster objecct.
```{r}
groups <- as.factor(cutree(clustG, k=6))
```

We will then join back the group list with shan_sf polygon feature data frame by using the code chun below.
```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

We can not plot the map of the newly delineated spatially constrained clusters.
```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```
